Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-25 icfp.tex
3.6.2 done

When the scrutinee is not in WHNF, it is not known statically which of the
scrutinee resources end up being consumed during evaluation and which would
still be available in the branch continuation.
%
Therefore, cases with non-WHNF scrutinees must by typed such that no resource
from the scrutinee can be used directly in the branch again, or risk
consuming the same resource twice.
%
However, simply evaluating the scrutinee to WHNF does not guarantee all
resources are consumed (unless the result matches an unrestricted pattern).
In the case where the matching pattern has some linear components

is not sufficient to
consume all resources used by the scrutinee, since sub-expressions such
as constructor arguments will be left unevaluated and may still contain
(``parts of'') the linear resources. To \emph{fully} consume all
resources occurring in the scrutinee, all linear components of the
constructor must also be fully evaluated, as witnessed by the
$\textsc{Alt}_0$ rule above.
%
In essence, to type cases scrutinizing expressions not in WHNF, we need to
forbid the usage of the scrutinee resources while tracking that we necessarily
``finish consuming'' them, either via the case binder, or by linearly using all
linear pattern-bound variables introduced in the alternative.

@@

We introduce \emph{proof irrelevant} resources, written as linear resources
within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$,
to encode linear resources that cannot be
directly used (i.e.,~to which the $\textsc{Var}$ rule is not applicable). Proof irrelevant resources
are linear resources in all other senses, meaning they must be used
exactly once. However, since proof irrelevant resources cannot be
discarded nor used directly, they have to be consumed \emph{indirectly} via
$\D$-bound variables, namely, the case binder, or the linear pattern-bound
variables.

Hence, to type a case expression whose scrutinee is not in WHNF, we
type the scrutinee with linear resources $\D$ and type the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, annotating the judgement with the proof irrelevant resources,
and using the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is similar to the one for scrutinees in WHNF, but
requiring the resources in the case binder, typing environment, and
judgement annotation to be made irrelevant.

@@

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Rightarrow_{\textsf{WHNF}}
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
% Essentially, if the scrutinee is in WHNF the matching alternative is easily
% determined, and must be treated with the specialized $\Mapsto$ judgement, which
% only applies to the matching constructor.
% %
Alternatives not matching the scrutinee could use resources arbitrarily as they
will never be executed, however, we uniformly treat non-matching alternatives
as if the scrutinee were not in WHNF. Having introduced proof irrelevant
resources, we can now present the correct $Case_\textrm{WHNF}$ rule:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely
written by humans.
Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
% Crucially, type preservation for the case-reduction substituting the
% case binder and pattern variables in the alternative is not possible
% to prove without branching on WHNF-ness, since otherwise the
% $\D$-substitution is not well-defined.  \todo[inline]{What is the
%   $\Delta$-substitution? Clarify}

@@

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
Linear Core is suitable as the intermediate language of an optimising compiler
for a linear and lazy language such as Haskell Core, in that optimising
transformations in Linear Core preserve types \emph{and} linearity, since Linear
Core understands semantic linearity in the presence of laziness, unlike
Core's current type system under which optimisations currently violate
linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
A linearly typed Core that preserves linearity throughout the optimisation
pipeline of GHC both validates the correctness of the compiler, which is
already achieved to a great extent by preserving (non-linear) types, and
informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must
account for all other Core features (e.g. strict constructor fields), propagate
new or modified data throughout the entire pipeline, and accept more
optimisations. Despite our initiative in this direction\footnote{[Link removed
due to
anonymization]}%\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
we leave this as future work.

>>>>

When the scrutinee is not in WHNF, it is not statically known which resources
will have been consumed after evaluating said scrutinee, or which
are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
a \rightarrow e\}}$ where $f = K$, but opaque.
\begin{enumerate}

    \item Consequently, cases with non-WHNF scrutinees must by typed such that no resource
    from the scrutinee can be used directly in the branch again, or risk
    consuming the same resource twice. In the example, $x$ must not directly occur
    in $e$ since $f$ \emph{could} have consumed it.

    \item On the other hand, since evaluation to WHNF does not guarantee all
    resources do end up being consumed (unless the result matches an
    unrestricted pattern), we must ensure any which transfer from the
    scrutinee to the alternative are ultimately consumed. (In the example,
    $x$ is not consumed by evaluating the scrutinee).

    Considering (1), we must resort to a more uniform method of guaranteeing we
    ``finish consuming`` the scrutinee. And there are only two ways of
    uniformly referring to the remaining linear resources in the newly-evaluated scrutinee:

        \begin{itemize}
            \item The case binder, which refers to the whole result of evaluating the scrutinee
            \item The linear components of a constructor pattern, which refer to expressions that may contain linear resources.
        \end{itemize}

%To guarantee all post-evaluation unconsumed scrutinee resources are consumed in
%the alternative, without statically knowing
%Statically we can't know which resources are ``transferred'', but  guarantee
%
\end{enumerate}

%is not sufficient to
%consume all resources used by the scrutinee, since sub-expressions such
%as constructor arguments will be left unevaluated and may still contain
%(``parts of'') the linear resources. To \emph{fully} consume all
%resources occurring in the scrutinee, all linear components of the
%constructor must also be fully evaluated, as witnessed by the
%$\textsc{Alt}_0$ rule above.
%%
%In essence, to type cases scrutinizing expressions not in WHNF, we need to
%forbid the usage of the scrutinee resources while tracking that we necessarily
%``finish consuming'' them, either via the case binder, or by linearly using all
%linear pattern-bound variables introduced in the alternative.

@@

We introduce \emph{proof irrelevant} resources, written as linear resources
within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$,
to encode linear resources that cannot be
directly used (i.e.,~to which the $\textsc{Var}$ rule is not applicable). Proof irrelevant resources
are linear resources in all other senses, meaning they must be used
exactly once. However, since proof irrelevant resources cannot be
discarded nor used directly, they have to be consumed \emph{indirectly} via
$\D$-bound variables, namely, the case binder, or, in mutual exclusion, the linear pattern-bound
variables.

Hence, to type a case expression whose scrutinee is not in WHNF, we
type the scrutinee with linear resources $\D$ and type the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, and annotating the judgement with the proof irrelevant resources
for use in the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is similar to the one for scrutinees in WHNF, but
requires the resources in the case binder, typing environment, and
judgement annotation to be made ``irrelevant''.

@@

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Rightarrow_{\textsf{WHNF}}
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
% Essentially, if the scrutinee is in WHNF the matching alternative is easily
% determined, and must be treated with the specialized $\Mapsto$ judgement, which
% only applies to the matching constructor.
% %
Alternatives not matching the scrutinee could use resources arbitrarily as they
will never be executed, however, we uniformly treat non-matching alternatives
as if the scrutinee were not in WHNF. Having introduced proof irrelevant
resources, we can now present the final $Case_\textrm{WHNF}$ rule:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely
written by humans.
Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
% Crucially, type preservation for the case-reduction substituting the
% case binder and pattern variables in the alternative is not possible
% to prove without branching on WHNF-ness, since otherwise the
% $\D$-substitution is not well-defined.  \todo[inline]{What is the
%   $\Delta$-substitution? Clarify}

@@

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
Linear Core is suitable as the intermediate language of an optimising compiler
for a linear and lazy language such as Haskell Core, in that optimising
transformations in Linear Core preserve types \emph{and} linearity, since Linear
Core understands semantic linearity in the presence of laziness, unlike
Core's current type system under which optimisations currently violate
linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
A linearly typed Core that preserves linearity throughout the optimisation
pipeline of GHC both validates the correctness of the compiler, which is
already achieved to a great extent by preserving (non-linear) types, and
informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must
account for all other Core features (e.g. strict constructor fields), propagate
new or modified data throughout the entire pipeline, and accept more
optimisations. Despite our initiative in this direction\footnote{[Link removed
due to
anonymization]} %\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
we leave this as future work.
